----
title: 'Reproducible Research - Assignment #2'
author: "seddonma"
date: "July 25, 2015"
---

# Population and Economic Costs of Severe Weather Events

## Synopsis 

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern. This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. The latest link of the dataset's documentation can be found here: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf. 

The research questions for this study are as follows:

1.  Across the United States, which types of events are most harmful with respect to population health?

2.  Across the United States, which types of events have the greatest economic consequences?

## Data Processing    
    
The data was most recently downloaded in July 25th, 2015. The file downloaded is a comma-separated-value file compressed via the bzip2 algorithm. Once the dataset is downloaded, it is loaded into memory in a dataframe called "dataset". 
```{r}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destination <- "repdata-data-StormData.csv.bz2"
setInternet2(use = TRUE)
download.file(url, destination)
dataset <- read.csv("repdata-data-StormData.csv.bz2", sep = ",", header = TRUE)
```

This data analysis requires the use of the plyr and ggplot2 packages. The following code snippet loads the two packages.
```{r}
library(plyr)
library(ggplot2)
library(stringr)
```

```{r}
str(dataset)
```

Using the dataset's documentation, an initial look at the full dataset shows a large number of variables that are irrelevant to the research questions, such as the location and time that each event occurred. Therefore these variables are removed. In addition, in order to optimize cleaning and analyzing the dataset, observations that do not list any population or economic damage are removed.
```{r}
dataset <- dataset[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
dataset <- subset(dataset, FATALITIES != 0 | INJURIES != 0 | PROPDMG != 0 | CROPDMG != 0)
```

In order to measure the impact of a weather event on the population's health, the number of injuries and fatalities that directly result from the weather event are summed into a variable called "POPDMG". 
```{r}
dataset$POPDMG <- dataset$INJURIES + dataset$FATALITIES
```

In order to determine the total economic cost of each weather event, many transformations needed to be made, largely due to the economic cost being divided into 4 columns. PROPDMG and CROPDMG are the dollar value of the property and crop damage resulting from the weather event. However, this dollar value does not include the magnitude of the dollar value, which is stored in the PROPDMGXP and CROPDMGXP variables and should be coded to H, K, M, and B for (respectively) a hundred, thousand, million, and billion dollars. 

First, all characters that do not denote a magnitude of money are removed. Secondly, because some characters are lowercase and others are uppercase, all remaining characters are capitalized to ensure consistency. Next, H, K, M, and B are recoded to their respective numeric equivilents. Then all empty cells are recoded to 0 to avoid "NA"s populating the dataset. Finally, the total property and crop variables are summed.
```{r}
dataset <- dataset[!(dataset$PROPDMGEXP %in% c("-", "+", "?", "0", "2", "3", "4", "5", "6", "7", "8")),]
dataset <- dataset[!(dataset$CROPDMGEXP %in% c("?", "0")),]
dataset$PROPDMGEXP <- toupper(dataset$PROPDMGEXP)
dataset$CROPDMGEXP <- toupper(dataset$CROPDMGEXP)
dataset$PROPDMGEXP[dataset$PROPDMGEXP == "H"] <- 100 
dataset$PROPDMGEXP[dataset$PROPDMGEXP == "K"] <- 1000
dataset$PROPDMGEXP[dataset$PROPDMGEXP == "M"] <- 1000000
dataset$PROPDMGEXP[dataset$PROPDMGEXP == "B"] <- 1000000000
dataset$CROPDMGEXP[dataset$CROPDMGEXP == "H"] <- 100 
dataset$CROPDMGEXP[dataset$CROPDMGEXP == "K"] <- 1000
dataset$CROPDMGEXP[dataset$CROPDMGEXP == "M"] <- 1000000
dataset$CROPDMGEXP[dataset$CROPDMGEXP == "B"] <- 1000000000
dataset$PROPDMGEXP <- sub("^$", 0, dataset$PROPDMGEXP)
dataset$CROPDMGEXP <- sub("^$", 0, dataset$CROPDMGEXP)
dataset$ECODMG <- (dataset$PROPDMG*as.numeric(dataset$PROPDMGEXP)) + (dataset$CROPDMG*as.numeric(dataset$CROPDMGEXP))
```

There is some duplication present in the EVTYPE variable, this is somewhat allivated by capitalizing all EVTYPE observations. 
```{r}
dataset$EVTYPE <- toupper(dataset$EVTYPE)
```

A look at the frequencies of the EVTYPE variable shows a considerable number of spelling and entry errors. Those errors with a frequency higher than 100 are recoded to match the dataset documentation's Storm Data Event Table found on page 6. The observations that pertain to two different kinds of weather events are not recoded in order to not bias one event over the other. 
```{r}
events_freq <- cbind(table(dataset$EVTYPE))
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "TSTM WIND", "THUNDERSTORM WIND")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "THUNDERSTORM WINDS", "THUNDERSTORM WIND")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "URBAN/SML STREAM FLD", "FLOOD")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "WILD/FOREST FIRE", "WILDFIRE")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "FLASH FLOODING", "FLASH FLOOD")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "RIP CURRENTS", "RIP CURRENT")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "HEAT", "EXCESSIVE HEAT")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "EXTREME COLD", "EXTREME COLD/WIND CHILL")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "LANDSLIDE", "DEBRIS FLOW")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "STORM SURGE", "STORM SURGE/TIDE")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "LIGHT SNOW", "WINTER WEATHER")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "URBAN FLOOD", "FLOOD")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "WINTER WEATHER/MIX", "WINTER WEATHER")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "MARINE TSTM WIND", "MARINE THUNDERSTORM WIND")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "FOG", "DENSE FOG")
dataset$EVTYPE <- str_replace_all(dataset$EVTYPE, "RIVER FLOOD", "FLOOD")
```

## Results


```{r}
dataset_pop <- ddply(dataset, "EVTYPE", summarize, POPDMG_sum = sum(POPDMG))
dataset_pop <- subset(dataset_pop, POPDMG_sum > 5000)
qplot(x = EVTYPE, y = POPDMG_sum, data = dataset_pop, geom = 'bar', stat = "identity", position = "stack") + labs(title = "Total Deaths and Injuries Per Severe Weather Event", x = "Severe Weather Event", y = "Sum of Deaths and Fatalities")
```


```{r}
dataset_econ <- ddply(dataset, "EVTYPE", summarize, ECODMG_sum = sum(ECODMG))
dataset_econ <- subset(dataset_econ, ECODMG_sum > 15000000000)
qplot(x = EVTYPE, y = ECODMG_sum, data = dataset_econ, geom = 'bar', stat = "identity") + labs(title = "Total Economic Cost Per Severe Weather Event", x = "Severe Weather Event", y = "Total Economic Cost")
```

Do the figure(s) have descriptive captions (i.e. there is a description near the figure of what is happening in the figure)?

EXPLAIN RESULTS, ANSWER ABOVE QUESTION AND AND YOU`RE DONE
